---
title: "Case Study V Interim Report I"
author: "Jake Epstein, Daniel Spottiswood, Sahil Patel, Michael Tan, Man-Lin Hsiao"
date: "11/10/2019"
output:
  pdf_document: default
  html_document: default
---

```{r, echo = FALSE, message = FALSE}
library(haven)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(xtable)
library(xgboost)
library(randomForest)
library(twang)
library(VIM)
library(lattice)
library(mice)
library(survival)
library(survminer)



cardio = read.csv("data/dukecathr.csv")
 
# CATEGORICAL OR BINARY VARIABLES:
cardio$DEATH = as.factor(cardio$DEATH)
cardio$GENDER = as.factor(cardio$GENDER)


select = dplyr::select
```

```{r}
ggplot(cardio, aes(x = DAYS2LKA, fill = DEATH==1)) +
  geom_density(alpha = 0.5) +
  labs(title = "Days to Death or Last Contact", fill = "Dead", x = "Days to Death or Last Contact", y = "Density") +
  theme_minimal()
```

This chart confirms what we expect. We see that most of the early deaths are accounted for but that longer term survivors often lose contact before a confirmed death.

```{r}
cardio_na = sapply(cardio, function(x) sum(is.na(x)))
cardio_nas = data.frame(cardio_na, colnames(cardio), stringsAsFactors = FALSE) %>% 
  mutate(cardio_na_p = 100*cardio_na/nrow(cardio)) %>%
  arrange(cardio_na) %>% filter(cardio_na > 0) 
cardio_nas$colnames.cardio. = factor(cardio_nas$colnames.cardio., levels = cardio_nas$colnames.cardio.)
ggplot(cardio_nas, aes(x = colnames.cardio., y = cardio_na_p, fill = colnames.cardio.)) +
  geom_bar(stat="identity", show.legend = FALSE) + 
  labs(title = "Missing Data Proportions", x = "", y = "Percent Missing") +
  theme_minimal() +
  coord_flip()
```

We see several variables with missing data rates over 50 percent, and many more in the ~20-40% range. 
```{r,echo=FALSE}
# md.pattern(cardio %>% dplyr::select(DSSTROKE, DSMI))
# cardio_aggr <- aggr(cardio,col=mdc(1:2),numbers=TRUE,sortVars=TRUE,
# labels=names(cardio),cex.axis=.7, gap=3, 
# ylab=c("Proportion missing","Missingness Pattern"))
```

Most instances of large missing data are for events that likely never occurred, such as all the "days to first subsequent" variables like DSSTROKE (having a stroke in the future) and DSMI (having a non-fatal myocardianl infarction in the future), or "days to closest previous" variables like DPCABG (having a previous coronary artery bypass surgery).

On the other hand, some variables with a significant amount of missing data do not have easily interpretable explanations. For example laboratory measurements such as LDL_R, HDL_R, and TOTCHOL_R have significant proportions of missing data, but there is no straightforward explanation to why these measurements may not have been recorded for certain patients.  


```{r}
deathdf = data.frame()
missingvals = list('DSSTROKE', 'DSMI', 'GRAFTST', 'DPCABG', 'DSCABG', 'DPPCI', 'LDL_R', 'HDL_R', 'TOTCHOL_R', 'DPMI', 'DSPCI', 'LVEF_R', 'CREATININE_R', 'LCXST', 'RCAST', 'PRXLADST', 'LADST', 'LMST', 'DIASBP_R', 'SYSBP_R', 'NUMDZV', 'CHFSEV', 'PULSE_R', 'S3')
for(val in missingvals){
  missing <- cardio[which(is.na(eval(parse(text=paste('cardio',"$",val,sep=""))))),]
  notmissing <- cardio[which(!is.na(eval(parse(text=paste('cardio',"$",val,sep=""))))),]
  
  percentdeathmissing = summary(missing$DEATH)[2] / (summary(missing$DEATH)[2] + summary(missing$DEATH)[1])
  percentdeathnotmissing = summary(notmissing$DEATH)[2] / (summary(notmissing$DEATH)[2] + summary(notmissing$DEATH)[1])
  
  deathdf[paste(val,'missing'), 'deathpercentage'] = percentdeathmissing
  deathdf[paste(val,'missing'), 'subset'] = 'missing'
  deathdf[paste(val, 'missing'), 'name'] = val
  deathdf[paste(val,'observed'), 'deathpercentage'] = percentdeathnotmissing
  deathdf[paste(val,'observed'), 'subset'] = 'observed'
  deathdf[paste(val, 'observed'), 'name'] = val
}

ggplot(deathdf, aes(name, deathpercentage, fill = subset)) + 
    geom_bar(stat = "identity", position = "dodge") + 
    labs(title="Percentage Deaths in Missing VS Observed Data", 
         x = "Variable", y = "% Deaths") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In order to investigate how missing data affects the outcome of death, we plotted the percentage of deaths in every variable for the subset where data on the variable is missing, and the subset where data on the variable is observed. Through the graph, we can notice that five specific variables: CHFSEV, CREATININE_R, PULSE_R, S3, and TOTCHOL_R have particularly large differences in percentage deaths for their observed versus missing subsets. For all five variables, there was a higher percentage of deaths for the subset of data that is missing. This suggests that for these variables, there is a possibility that data may be MNAR (missing not at random). 


```{r}
marginplot(cardio[, c("CREATININE_R", "DAYS2LKA")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
marginplot(cardio[, c("S3", "DAYS2LKA")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)

marginplot(cardio[, c("PULSE_R", "DAYS2LKA")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
```
We continued to investigate the five aforementioned variables and their effects on DAYS2LKA through margin plots. Although there were no noteworthy differences in distribution for CHSEV and TOTCHOL_R, we did find that missing data seems to effect DAYS2LKA for the variables CREATININE_R, S3, and PULSE_R. For all three of these variables, subsets of missing data have boxplots with higher means, indicating that patients missing data on these variables died later than patients with observed data. 

We will proceed with visualizations that may help determine any similar patterns for missing data in different variables.

```{r}
cardio_aggr <- aggr(cardio[,c(38:41)],col=mdc(1:2),numbers=TRUE,sortVars=TRUE,
labels=names(cardio[,c(38:41)]),cex.axis=.7, gap=3, 
ylab=c("Proportion missing","Missingness Pattern"))

cardio_aggr <- aggr(cardio[,c(30:33)],col=mdc(1:2),numbers=TRUE,sortVars=TRUE,
labels=names(cardio[,c(30:33)]),cex.axis=.6, gap=3, 
ylab=c("Proportion missing","Missingness Pattern"))
```

Through examining the variables with the highest proportion of missing data with missingness pattern visualizations, the following can be seen:

1) When looking at GRAFTST, LCXST, LADST, and LMST, all four of these have missing values together 10.4% of the time. These are all catetherization results, so it could be the case the post-catetherization measurements were not done well or completely. It is likely that the missing data results from cases where the doctor just did not bother to measure all the metrics after catetherization, which would mean that the data is missing completely at random (MCAR).

2) When looking at LDL_R, HDL_R, TOTCHOL_R, and CREATININE_R, the variables excluding CREATININE_R have missing values together 37% of the time, and all four are missing 24% of the time. We continue to investigate this in the following graphs.

```{r}
marginplot(cardio[, c("CREATININE_R", "YRCATH_G")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
marginplot(cardio[, c("TOTCHOL_R", "YRCATH_G")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
marginplot(cardio[, c("HDL_R", "YRCATH_G")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
marginplot(cardio[, c("LDL_R", "YRCATH_G")], col = mdc(1:2), cex.numbers = 0.4, pch = 1)
```

Through these margin plots, we can see there is a significant non-overlap in the distribution of YRCATH_G (year of catetherization) data depending on whether the laboratory measurements were missing versus observed. The majority of missing data on all these laboratory measurements comes from early years, while all the observed data comes from later years. A likely explanation is that these laboratory procedures were less common or accessible in the earlier years. From this, we can determine that this data is missing at random (MAR). 


Through these exploratory data analysis and missing data visualizations, we can determine that significant portions of missing data are due to either events that never occurred and therefore cannot be recorded, possible negligence in recording certain measurements (MCAR), and reasons that depend on other observed variables, such as less accessible laboratory procedures in earlier years leading to missing data (MAR).


### Build Model

In order to address the issues associated with missing data, we built a complete case, and estimated survival functions using both Kaplan-Meier curves and Cox proportional hazards regression modelling.

[TODO: add more variables here to explore]

```{r, fig.height= 15, fig.width = 7}
# select data we want
# note: selected from fan's slides

cardio_new = cardio %>%
  select(
    YRCATH_G, #year
    AGE_G, #age
    GENDER,
    RACE_G,
    HXSMOKE, #history of smoking
    WEIGHT_R, # in kg
    HEIGHT_R, # in cm
    HXCEREB, #history of stroke
    CHFSEV, # congestive heart failure severity
    HXCOPD, # history of COPD
    HXDIAB, # diabetes 
    DPMI, #days to previous myocardial infarction. No MI encoded as missing --> will change this to a binary 
    HXHYL, #history of hyperlipidemia 
    S3, #S3 gallop
    LMST, #maximum stenosis of left main artery
    NUMDZV, #number of diseased vessels
    DEATH,
    DAYS2LKA, # days until last known alive
    CREATININE_R, #most recent serum creatinine
    TOTCHOL_R, #total cholesterol
    HDL_R, #high density lipid
    LDL_R #low density lipid
  ) %>%
  mutate(BMI = WEIGHT_R / (HEIGHT_R / 100) ^ 2, HXMI = 1 - is.na(DPMI), DEATH = as.numeric(DEATH), RACE_G = as.factor(RACE_G),
         CHFSEV = as.factor(CHFSEV), LMST = 100 * LMST, YRCATH_G = as.factor(YRCATH_G), AGE_G = as.factor(AGE_G)) %>%
  select(-WEIGHT_R, -HEIGHT_R, -DPMI)
```

The plots below depict the Kaplan-Meier curves (the risk of death over time) of variables we found to demonstrate clear effects that we could draw causal inference from. The plots also include the p-value dervied from conducting log-rank tests on the Kaplan-Meier curves.
</br>
</br>
The Kaplan-Meier curves are depictions of the variable's survival function over time. The probability of survival begins at 1 (time = 0), and approaches 0 as time goes to $\infty$. The log-rank test determines whether the differences between a variable's Kaplan-Meier curves are statistically significant. All of the variables examined had p-values <0.0001, implying that they were statistically significant.

```{r}
#Kaplan-Meier curves -- tool to decide what to put into the model we use

#gender
fit_gender=survfit(Surv(DAYS2LKA,DEATH)~GENDER,data=cardio_new)
ggsurvplot(fit_gender,data=cardio_new,pval=TRUE)

#age (low is <40)
#fit_age=survfit(Surv(DAYS2LKA,DEATH)~AGE_G,data=cardio_new)
#ggsurvplot(fit_age,data=cardio_new,pval=TRUE)

cardio_new=cardio_new%>%
  mutate(age_low=case_when(AGE_G%in%seq(1,4,1)~1,AGE_G%in%seq(4,13,1)~0))
fit_age_low=survfit(Surv(DAYS2LKA,DEATH)~age_low,data=cardio_new)
ggsurvplot(fit_age_low,data=cardio_new,pval=TRUE)

#race
#fit_race=survfit(Surv(DAYS2LKA,DEATH)~RACE_G,data=cardio_new)
#ggsurvplot(fit_race,data=cardio_new,pval=TRUE)

#CHF severity
fit_sev=survfit(Surv(DAYS2LKA,DEATH)~CHFSEV,data=cardio_new)
ggsurvplot(fit_sev,data=cardio_new,pval=TRUE)

#year of catheter
#fit_yr=survfit(Surv(DAYS2LKA,DEATH)~YRCATH_G,data=cardio_new)
#ggsurvplot(fit_yr,data=cardio_new,pval=TRUE)

#serum creatinine
cardio_new=cardio_new%>%
  mutate(creatinine_levels=case_when(
    CREATININE_R<0.9~1,
    CREATININE_R>=0.9&CREATININE_R<1~2,
    CREATININE_R>=1&CREATININE_R<1.3~3,
    CREATININE_R>=1.3~4,
  ))
fit_serum=survfit(Surv(DAYS2LKA,DEATH)~creatinine_levels,data=cardio_new)
ggsurvplot(fit_serum,data=cardio_new,pval=TRUE)

#total cholesterol
cardio_new=cardio_new%>%
  mutate(chol_levels=case_when(
    TOTCHOL_R<146~1,
    TOTCHOL_R>=146&TOTCHOL_R<174~2,
    TOTCHOL_R>=174&TOTCHOL_R<205~3,
    TOTCHOL_R>=205~4,
  ))
fit_chol=survfit(Surv(DAYS2LKA,DEATH)~chol_levels,data=cardio_new)
ggsurvplot(fit_chol,data=cardio_new,pval=TRUE)

#high denisty lipid
cardio_new=cardio_new%>%
 mutate(hdl_levels=case_when(
   HDL_R<33~1,
   HDL_R>=33&HDL_R<39~2,
   HDL_R>=39&HDL_R<48~3,
   HDL_R>=48~4,
 ))
#fit_hdl=survfit(Surv(DAYS2LKA,DEATH)~hdl_levels,data=cardio_new)
#ggsurvplot(fit_hdl,data=cardio_new,pval=TRUE)

#low denisty lipid
cardio_new=cardio_new%>%
  mutate(ldl_levels=case_when(
    LDL_R<77~1,
    LDL_R>=77&LDL_R<100~2,
    LDL_R>=100&LDL_R<127~3,
    LDL_R>=127~4,
  ))
fit_ldl=survfit(Surv(DAYS2LKA,DEATH)~ldl_levels,data=cardio_new)
ggsurvplot(fit_ldl,data=cardio_new,pval=TRUE)
```

The Kaplan-Meier curves for the variable gender shows that females have a statistically significant higher risk of dying than males. In addition, the curves show that this difference in risk of death increases as time increases (the slope of the red, male, curve increases faster than the blue, female curve). 
</br>
</br>
In order to determine whether there was an effect for age, we categorised an individual's age to  be low ($<$40) or high ($\geq$40). The Kaplan-Meier curves for this variable show a distinguished difference in the risk of death for the two groups. Namely, individuals who are 40 or older have a statisitcally significant increased risk of death. Furthermore, as time passes, this difference increases furhter; in other words, as time passes individuals who are 40 or older have an increasing risk of dying when compared to those under the age of 40. Contextually, this is a reasonable outcome as individuals who are older are more likely to die than their juniors.
</br>
</br>
The Kaplan-Meier curves for congestive heart failure (CHF) severity indicate that individuals who never experienced CHF have the least risk of dying than those have had CHF. Moreover, the risk of dying increases as CHF severity increases. Intuitively, this is a feasible outcome as individuals with the most severe CHF are characterised by an "inability to perform any physical activity without discomfort" and potential discomfort while resting; logically, individuals who experience this level of pain are at higher risk of dying than those with less severe symptoms.
</br>
</br>
creatinine
</br>
</br>
cholesterol
</br>
</br>
ldl
</br>
</br>
We will explore these relationships further using Cox proportional hazards regression modelling. Hopefully, the Cox regression methodology will provide more insight into how these differences in risk of dying manifest at a more granular level.

[TO DO: finish write up^]

```{r}
cardio_new = cardio_new %>%
  mutate(chol_levels = as.factor(chol_levels)) %>%
  mutate(hdl_levels = as.factor(hdl_levels)) %>%
  mutate(ldl_levels = as.factor(ldl_levels))

#complete case model
formula_call = Surv(DAYS2LKA, DEATH) ~ YRCATH_G + AGE_G + GENDER + RACE_G + HXSMOKE + HXCEREB + CHFSEV + HXCOPD + HXDIAB + HXHYL + S3 + LMST + NUMDZV + creatinine_levels + chol_levels + hdl_levels + ldl_levels #note : all vars with "_levels" at the end are categorised by quantile

fit.coxph <- coxph(formula_call, data = cardio_new)

ggsave("output/cox_graph_complete_case.png", ggforest(fit.coxph, data=cardio_new), device = "png", width = 7, height = 10, units = "in")
```

The forest plot above illustrates the hazard ratios for different values of the covariates explored as estimated through Cox proportional hazards regression modelling. A hazard ratio >1 implies that individuals with the associated value for the covariate have an increased chance of dying. Intuitively, hazard ratios <1 imply that the individual indicate a decreased risk of dying.
</br>
</br>
In order to run the Cox model, we established that our survival object would measure time with DAYS2LKA as we are working with censored, not interval, data, and we set the event to be 0 if the individual died and 1 if the individual stayed alive. We modelled the survival object using the following covariates: year of cardiac catheter (categorised), age in years (categorised), gender, race, history of smoking (binary), history of cerbrovascular disease (binary), congestive heart failure severity (binary), history of chronic obsructive pulmonary disease (binary), history of diabetes (binary), history of hyperlipidemia (binary), third hear sound (binary), maximum stenosis of left main artery, number of significantly diseased valves.
</br>
</br>
Our forest plot indicates that as an individual's age increases from 18-39 their risk of dying decreases; however, this trend reverses for individuals ages 39 and over. Furthermore, the same trend occurs for white and non-white individuals, where non-white individuals are at higher risk of dying. Moreover, this trend appears in the variable CHFSEV as well, where individuals with a severity of 0 are the least at risk. Interestingly, YRCATH_G has decreasing risk as the group number moves from 1-3, then increases for groups 4-7.


[TODO: give context to why the relationship do or don't make sense]

[TODO: justification for why we want to do MICE, why we chose the m we did, etc]


```{r}
# use multiple imputation with mice
#note: this takes a long time to run, so toggle rerunMICE to recalculate
#think we could make this go faster by parsing every factor as an int before running MICE, then refactoring before running the coxph
sum(cardio_new %>% is.na())/length(cardio_new)/nrow(cardio_new)
#about 20% of data is missing--> m should be 20


rerunMICE = FALSE

if(rerunMICE){
  imp = mice(cardio_new, m = 10)
  saveRDS(imp, "data/imp.rds")
} else {imp = readRDS("data/imp.rds")}


densityplot(imp)

#note, cant replace formula with formula_call here for some reason
coximpute = with(imp,coxph(Surv(DAYS2LKA, DEATH) ~ YRCATH_G + AGE_G + GENDER + RACE_G + HXSMOKE + HXCEREB + CHFSEV + HXCOPD + HXDIAB + HXHYL + S3 + LMST + NUMDZV))
summary(pool(coximpute))
```

```{r}
# compare methods

mice_output = summary(pool(coximpute)) %>% select(estimate, std.error)
mice_output$variable = rownames(mice_output)
mice_output$procedure = "MICE"

fixed_output = data.frame(estimate = summary(fit.coxph)$coefficients[,1],  std.error = summary(fit.coxph)$coefficients[,3])
fixed_output$variable = rownames(fixed_output)
fixed_output$procedure = "compete cases"

output = rbind(mice_output, fixed_output)

comparison_plot = ggplot(output, aes (x = variable, y = exp(estimate), color = procedure))+
  geom_point(shape = "square", size = 3)+
  geom_errorbar(aes(ymin = exp(estimate - 2 * std.error), ymax = exp(estimate + 2 * std.error))) +
  geom_hline(yintercept = 1) +
  coord_flip() +
  scale_y_continuous(trans = "log2")+
  labs(main = "Comparison of Procedures", x = "Variable", y = "Hazard Ratio", fill = "Procedure")+
  theme_minimal()


ggsave("output/cox_graph_compare.png", comparison_plot, device = "png", width = 7, height = 8, units = "in")

```

```{r}
#pred <- survfit(fit.coxph, cardio_new)
#timepoints <- c(0, pred$YR_CATH_)
```
